Annotation exploration for `r params$scpca_project_id`

### 2024-09-23 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
changes to PR#776

```{r packages, message=FALSE, warning=FALSE}
library(Seurat)
library(sctransform)
library(SCpubr)             # for plotting 
library(tidyverse)
library(patchwork)

library(reshape2)
library(org.Hs.eg.db)

@@

We work with the pre-processed and labeled `Seurat` object that is the output of `02b_label-transfer_fetal_kidney_reference_Stewart.Rmd` saved in the `results` directory.

@@

@@

metadata <- read.table(sample_metadata_file, sep = "\t", header = TRUE)
```

To explore the annotation results, we look into some marker genes reported here:

```{r marker_genes, fig.width=12, fig.height=6, out.width='100%'}
CellType_metadata <- read_csv(file.path(module_base, "marker-sets", "CellType_metadata.csv"))

DT::datatable(CellType_metadata, caption = ("CellType_metadata"), 
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',
                               buttons = c( 'csv', 'excel')))
```

@@

Here we defined function that will be used multiple time all along the notebook.

#### Visualize seurat clusters and metadata

For a Seurat object `object`and a metadata `metadata`, the function `visualize_metadata` will plot `FeaturePlot` and `BarPlot`

- `object` is the Seurat object

- `metadata` the gene or quantitative value to be plotted

- `group.by` is the metadata used for grouping the violin plots

```{r }
visualize_metadata <- function(object, meta, group.by){

  if(is.numeric(object@meta.data[,meta])){
     d <- SCpubr::do_FeaturePlot(object, 
                                              features = meta, 
                                              reduction = "umap.rpca",
                                              pt.size = 0.2, 
                                              legend.width = 0.5, 
                                              legend.length = 5, 
                                              legend.position = "right") + ggtitle(meta)
    b <- SCpubr::do_ViolinPlot(srat, 
                                             features = meta, 
                                             ncol = 1, 
                                             group.by = group.by, 
                                             legend.position = "none")

   return(d + b + plot_layout(ncol = 2, widths = c(2,4))) 
  }

  else{

  d <- SCpubr::do_DimPlot(object, reduction="umap.rpca", group.by = meta, label = TRUE, repel = TRUE) + ggtitle(paste0(meta," - umap")) + theme(text=element_text(size=18)) + NoLegend()
  b <- SCpubr::do_BarPlot(sample = object,
                         group.by = meta,
                         split.by = group.by,
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 3) +
                         ggtitle("% cells")+
                         xlab(print(group.by)) +
                         theme(text=element_text(size=18))
  return(d + b + plot_layout(ncol = 1, heights = c(4,2)))
  }

}
```

#### Visualize seurat clusters and markers genes

For a Seurat object `object`and a features `features`, the function `visualize_feature` will plot `FeaturePlot` and `ViolinPlot`

- `object` is the Seurat object

- `features` the gene or quantitative value to be plotted

- `group.by` is the metadata used for grouping the violin plots

```{r }
visualize_feature <- function(object, features, group.by = "seurat_clusters"){

                  feature_symbol <- AnnotationDbi::select(org.Hs.eg.db, 
                                keys=features, 
                                columns="SYMBOL", 
                                keytype="ENSEMBL")

                  d <- SCpubr::do_FeaturePlot(object, 
                                              features = feature_symbol$ENSEMBL,
                                              reduction = "umap.rpca",
                                              pt.size = 0.2, 
                                              legend.width = 0.5, 
                                              legend.length = 5, 
                                              legend.position = "right") + ggtitle(feature_symbol$SYMBOL)
                  b <- SCpubr::do_ViolinPlot(srat, 
                                             features = feature_symbol$ENSEMBL, 
                                             ncol = 1, 
                                             group.by = group.by, 
                                             legend.position = "none",
                                             assay = "SCT") + ylab(feature_symbol$SYMBOL)

                  return(d + b + plot_layout(ncol = 2, widths = c(2,4))) 
}

@@

- `data` seurat object 
- `first_group` is the name of the first metadata to group the cells
- `last_group` is the name of the second metadata to group the cells

```{r fig.width=10, fig.height=4, out.width='100%'}
do_Table_Heatmap <- function(data, first_group, last_group ){

  df <- data@meta.data %>%
  mutate_if(sapply(data@meta.data, is.character), as.factor)  %>% 
  group_by( !!sym(first_group), !!sym(last_group))%>% 
  summarise(Nb = log(n()))

p <- ggplot(df, aes(x= !!sym(first_group), y = !!sym(last_group), fill = Nb)) +
   geom_tile() +
   scale_fill_viridis_c()  +
   theme_bw() +
   theme(text = element_text(size = 20)) +
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
   guides(fill=guide_legend(title=" Number of \n cells (log)"))

@@

### Load the pre-processed `Seurat` object
```{r load, message=FALSE, warning=FALSE}
# open the processed rds object
filelist <- list.files(result_dir, 
                       recursive = TRUE, 
                       pattern = "02b-fetal_kidney_label-transfer_", 
                       full.names = TRUE)

sample_id <- list.files(result_dir, 
                       recursive = TRUE, 
                       pattern = "02b-fetal_kidney_label-transfer_", 
                       full.names = FALSE)
sample_id <- colsplit(sample_id, pattern = "/", names = c("sample_id", "suffixe"))
sample_id <- sample_id[,1]

sce <- lapply(filelist, function(x) readRDS(x))

### Merge `Seurat` objects (with integration)

We followed the standard `Seurat` workflow described here to merge and integrate the 40 samples from the Wilms tumor dataset `r params$scpca_project_id`.
For more information, see the vignette [Merge objects (with integration)](https://satijalab.org/seurat/articles/essential_commands.html#merge-objects-with-integration). We have slight changes as we already performed normalization using [`SCTransform`](https://github.com/satijalab/seurat/issues/2814).

```{r process, message=FALSE, warning=FALSE}
options(future.globals.maxSize= 8912896000000)

srat <- merge(sce[[1]], y = sce[2:length(sce)], 
              add.cell.ids = sample_id, 
              project = params$scpca_project_id)

DefaultAssay(object = srat) <- "SCT"

VariableFeatures(srat[["SCT"]]) <- rownames(srat[["SCT"]]@scale.data)

srat <- RunPCA(srat)

srat <- IntegrateLayers(object = srat, 
                        method = RPCAIntegration,
                        orig.reduction = "pca", 
                        new.reduction = 'rpca',
                        normalization.method="SCT", verbose = FALSE)

srat <- FindNeighbors(srat, dims = 1:30, reduction = "rpca")
srat <- FindClusters(srat)
srat <- RunUMAP(srat, reduction = "rpca", dims = 1:30, reduction.name = "umap.rpca")
```

### Quick look at the integration

This is not the main point of this notebook, we expect to work on integration later on, one of the final step of the analysis. 
We simply here would like to have a rough idea of cell/sample distribution.

```{r fig.width=15, fig.height=10, out.width='100%'}
tmp <- colsplit(colnames(srat), pattern = "_", names = c("sample_id", "barcode"))
srat <- AddMetaData(srat, tmp$sample_id, col.name = "sample_id")

d1 <- SCpubr::do_DimPlot(srat, reduction="umap.rpca", group.by = "sample_id", label = FALSE) + ggtitle("sample_id - umap.rpca reduction")

d2 <- SCpubr::do_DimPlot(srat, reduction="umap.rpca", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("seurat_clusters - umap.rpca reduction")

d1 | d2
```

### Cell annotation

#### Introduction to the four annotation strategies

The use of the right reference is crucial for label transfer and annotation. 
It is recommended that the cell types in the reference is representative to the cell types to be annotated in the query.

Wilms tumors can contain up to three histologies that resemble fetal kidney: blastema, stroma, and epithelia [[1](https://www.ncbi.nlm.nih.gov/books/NBK373356/), [2](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9915828/)]. 
Because of their histological similarity to fetal kidneys, Wilms tumors are thought to arise from developmental derangements in embryonic renal progenitors.

For these reasons, we expect Wilms tumor cells to map to fetal cells, especially fetal kidney cells.

Here, we investigate annotations resulting from 4 different strategies:

* the `_processed.rds` objects already contain automated annotations computed by the [Data Lab](https://scpca.readthedocs.io/en/latest/getting_started.html):

    + Annotations determined by SingleR, an automated reference-based method [Looney et al. 2019](https://www.nature.com/articles/s41590-018-0276-y).

    + Annotations determined by CellAssign, an automated marker-gene based method [Zhang et al. 2019](https://www.nature.com/articles/s41592-019-0529-1).

* the label transfers from two fetal references using `Azimuth`, that have been performed in the previous analysis

    + Using the human fetal atlas as a reference: 
    This is the reference developed by [Cao et al.](https://www.science.org/doi/10.1126/science.aba7721) provided by `Azimuth` for label transfer.
    The reference contain cells from 15 organs including kidney from fetal samples.
    The label transfer have been performed in the notebook `02a_fetal_full_reference_Cao_{sample_id}.Rmd`

    + Using the human fetal kidney atlas: 
    [Stewart et al.](https://www.science.org/doi/10.1126/science.aat5031) created a [human fetal kidney atlas](https://www.kidneycellatlas.org/). 
    This reference contains only fetal kidney cells and has been precisely annotated by kidney experts. 
    The label transfer have been performed in the notebook `02b_fetal_kidney_reference_Stewart_{sample_id}.Rmd`

#### Annotation from SingleR (no cancer or kidney specific dataset)

```{r fig.width=20, fig.height=20, out.width='100%'}
visualize_metadata(srat, meta = "singler_celltype_annotation", group.by = "seurat_clusters")
```

#### Annotation from cellassign (no cancer or kidney specific dataset)

```{r fig.width=10, fig.height=20, out.width='100%'}
visualize_metadata(srat, meta = "cellassign_celltype_annotation", group.by = "seurat_clusters")
```

#### Annotation from the fetal reference (Cao et al)

@@

```{r fig.width=15, fig.height=20, out.width='100%'}
visualize_metadata(srat, meta = "fetal_full_predicted.organ", group.by = "seurat_clusters")
```

```{r fig.width=30, fig.height=20, out.width='100%'}
tmp <- table(srat$fetal_full_predicted.organ)
percent.of.kidney <- tmp["Kidney"] / sum(tmp) * 100
```

Of note, `r percent.of.kidney` % of the cells are labeled as kidney cells.

```{r fig.width=15, fig.height=20, out.width='100%'}
visualize_metadata(srat, meta = "fetal_full_predicted.annotation.l1", group.by = "seurat_clusters")
visualize_metadata(srat, meta = "fetal_full_predicted.annotation.l2", group.by = "seurat_clusters")
```

#### Annotation from the kidney reference (Stewart et al)

@@

```{r fig.width=10, fig.height=10, out.width='100%'}
visualize_metadata(srat, meta = "fetal_kidney_predicted.compartment", group.by = "seurat_clusters")
```

```{r fig.width=30, fig.height=20, out.width='100%'}
tmp <- table(srat$fetal_kidney_predicted.compartment)
percent.of.immune <- 100*tmp["immune"]/sum(tmp)
percent.of.endothelial <- 100*tmp["endothelium"]/sum(tmp)
```

Of note, `r percent.of.immune` % of the cells are labeled as immune cells and `r percent.of.endothelial` % of the cells are labeled as endothelial cells.

Let's have a look at the repartition of cells per patient:

```{r fig.width=10, fig.height=50, out.width='100%'}
table(srat$sample_id, srat$fetal_kidney_predicted.compartment)
```

```{r fig.width=10, fig.height=20, out.width='100%'}
visualize_metadata(srat, meta = "fetal_kidney_predicted.cell_type", group.by = "seurat_clusters")
```

### Compare annotations with marker genes

#### "fetal_kidney_predicted.compartment"

Here we evaluate with marker genes the identification of endothelial, immune cells and other cancer and non-cancerous sub-populations.

```{r fig.height=8, fig.width=8, message=FALSE, warning=FALSE, out.width='100%'}

markers = list("malignant" = CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "malignant")],
            "immune"= CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "immune")],
            "endothelium"= CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "endothelium")],
            "non-malignant"= CellType_metadata$ENSEMBL_ID[which(CellType_metadata$cell_class == "non-malignant")])

SCpubr::do_DotPlot(sample = srat,
                         group.by =  "fetal_kidney_predicted.compartment",
                         features = markers, 
                         axis.text.x.angle = 90, 
                         flip = TRUE, 
                         plot.title = "fetal_kidney_predicted.compartment"
                         ) 
```

@@

```{r fig.height=4, fig.width=8, message=FALSE, warning=FALSE, out.width='100%'}
visualize_feature(object = srat,
                  features = "ENSG00000110799",
                  group.by = "fetal_kidney_predicted.compartment")
```

@@

```{r fig.height=4, fig.width=8, message=FALSE, warning=FALSE, out.width='100%'}
visualize_feature(object = srat,
                  features = "ENSG00000081237",
                  group.by = "fetal_kidney_predicted.compartment")
```
#### "fetal_kidney_predicted.cell_type"

```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE, out.width='100%'}

SCpubr::do_DotPlot(sample = srat,
                         group.by =  "fetal_kidney_predicted.cell_type",
                         features = markers, 
                         axis.text.x.angle = 90, 
                         flip = TRUE,
                         plot.title = "fetal_kidney_predicted.cell_type"
                          )

```
#### "fetal_full_predicted.annotation.l1"

```{r fig.height=10, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}

SCpubr::do_DotPlot(sample = srat,
                         group.by =   "fetal_full_predicted.annotation.l1",
                         features = markers, 
                         axis.text.x.angle = 90, 
                         flip = TRUE,
                         plot.title =  "fetal_full_predicted.annotation.l1"
                          )
```
#### "cellassign_celltype_annotation"

```{r fig.height=15, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
SCpubr::do_DotPlot(sample = srat,
                         group.by =  "cellassign_celltype_annotation",
                         features = markers, 
                         axis.text.x.angle = 90, 
                         flip = TRUE,
                         plot.title = "cellassign_celltype_annotation"
                          )                         
```

### Compare annotations between them

#### Annotation versus `Seurat` clusters

Here we check wether one (or more) `Seurat` cluster do correspond to one label. 
This would allow us to be more confident in the annotation of immune or endothelial cells for example.

```{r fig.width=25, fig.height=25, out.width='100%'}
p1 <- do_Table_Heatmap( srat,
                            last_group = "fetal_kidney_predicted.compartment",
                            first_group = "seurat_clusters"

                            )

p2 <- do_Table_Heatmap( srat,
                            last_group = "fetal_full_predicted.annotation.l1",
                            first_group = "seurat_clusters"
                            )

p3 <- do_Table_Heatmap( srat,
                            last_group = "singler_celltype_annotation",
                            first_group = "seurat_clusters"
                            )

p4 <- do_Table_Heatmap( srat,
                            last_group = "cellassign_celltype_annotation",
                            first_group = "seurat_clusters"
                            )

(p1|p2)/(p3|p4)
```

#### Comparison of the labels from the two fetal references

##### first level annotation

```{r fig.width=30, fig.height=12, out.width='100%'}
p1 <- do_Table_Heatmap( srat,
                            last_group = "fetal_kidney_predicted.compartment",
                            first_group = "fetal_full_predicted.annotation.l1"

                            )
p2 <- do_Table_Heatmap( srat,
                            last_group = "fetal_kidney_predicted.cell_type",
                            first_group = "fetal_full_predicted.annotation.l1"

                            )
p1|p2
```

##### second level annotation

```{r fig.width=30, fig.height=15, out.width='100%'}
p2 <- do_Table_Heatmap( srat,
                            last_group = "fetal_kidney_predicted.cell_type",
                            first_group = "fetal_full_predicted.annotation.l2"

                            )
p2
```

#### Comparison of the  fetal reference with `singleR` and `cellassign`

##### first level annotation

```{r fig.width=30, fig.height=10, out.width='100%'}
p1 <- do_Table_Heatmap( srat,
                            last_group = "fetal_kidney_predicted.compartment",
                            first_group = "cellassign_celltype_annotation"

                            )
p2 <- do_Table_Heatmap( srat,
                            last_group = "fetal_kidney_predicted.compartment",
                            first_group = "singler_celltype_annotation"

                            )
p1|p2
```

##### second level annotation

```{r fig.width=30, fig.height=30, out.width='100%'}
p1 <- do_Table_Heatmap( srat,
                            last_group = "fetal_kidney_predicted.cell_type",
                            first_group = "cellassign_celltype_annotation"

                            )

p2 <- do_Table_Heatmap( srat,
                            last_group = "fetal_kidney_predicted.cell_type",
                            first_group = "singler_celltype_annotation"

                            )

p1/p2
```

>>>>

```{r packages, message=FALSE, warning=FALSE}
library(Seurat)
library(tidyverse)
library(patchwork)

@@

@@

We extracted from the pre-processed and labeled `Seurat` object (that is the output of `02b_label-transfer_fetal_kidney_reference_Stewart.Rmd` saved in the `results` directory)
the following information per cell:

- the predicted compartment in `fetal_kidney_predicted.compartment`

- the predicted organ in `fetal_full_predicted.organ`

- the sample identifier in `sample_id`

- the `counts`of the endothelial marker VWF = "ENSG00000110799"

- the `counts`of the immune marker PTPRC = "ENSG00000081237"

@@

metadata <- read.table(sample_metadata_file, sep = "\t", header = TRUE)

# Create a data frames of all annotations
cell_type_df <- metadata$scpca_sample_id |>
  purrr::map(
    # For each sample_id, do the following:
    \(sample_id) {

      # Read in the Seurat object
      srat <- readRDS(
        file.path(result_dir, sample_id, paste0("02b-fetal_kidney_label-transfer_", sample_id, ".Rds"))
      )

      # Create and return a data frame from the Seurat object with relevant annotations
      # this data frame will have four columns: barcode, sample_id, compartment, organ
      data.frame(
        compartment = srat$fetal_kidney_predicted.compartment,
        organ = srat$fetal_full_predicted.organ, 
        PTPRC = FetchData(object = srat, vars = "ENSG00000081237", layer = "counts"),
        VWF = FetchData(object = srat, vars = "ENSG00000110799", layer = "counts")
        ) |>
        tibble::rownames_to_column("barcode") |>
        dplyr::mutate(sample_id = sample_id)
    }
  ) |>
  # now combine all dataframes to make one big one
  dplyr::bind_rows()

@@

@@

- `first_group` is the name of the first metadata to group the cells
- `last_group` is the name of the second metadata to group the cells

```{r fig.width=10, fig.height=4, out.width='100%'}
do_Table_Heatmap <- function(first_group, last_group ){

  df <- cell_type_df %>%
  group_by( !!sym(first_group), !!sym(last_group))%>% 
  summarise(Nb = n())  %>%
  mutate(C = sum(Nb)) %>%
  mutate(percent = Nb/C*100)

p1 <- ggplot(df, aes(x= !!sym(first_group), y = !!sym(last_group), fill = Nb)) +
   geom_tile() +
   scale_fill_viridis_c()  +
   theme_bw() +
   theme(text = element_text(size = 20)) +
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
   guides(fill=guide_legend(title=" Number of \n cells"))

p2 <- ggplot(df, aes(x= !!sym(first_group), y = !!sym(last_group), fill = percent)) +
   geom_tile() +
   scale_fill_viridis_c()  +
   theme_bw() +
   theme(text = element_text(size = 20)) +
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
   guides(fill=guide_legend(title=" Percentage of \n cells"))

p <- p1 + p2 + plot_layout(ncol = 2)
return(p)
}
```

#### do_Feature_mean

`do_Feature_mean` shows heatmap of mean expression of a feature grouped by a metadata.

- `group.by` is the name of the metadata to group the cells
- `feature` is the name of the gene to average the expression

```{r fig.width=10, fig.height=4, out.width='100%'}
do_Feature_mean <- function(group.by, feature ){

  df <- cell_type_df %>%
  group_by(sample_id, !!sym(group.by))%>% 
  summarise(m = mean(!!sym(feature)))

p <- ggplot(df, aes(x= sample_id, y = !!sym(group.by), fill = m)) +
   geom_tile() +
   scale_fill_viridis_c()  +
   theme_bw() +
   theme(text = element_text(size = 20)) +
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
   guides(fill=guide_legend(title=paste0(feature)))

@@

### Total number of cells per samples

```{r fig.width=5, fig.height=10, out.width='100%', results='asis'}
tmp <- table(cell_type_df$sample_id)
knitr::kable(tmp)
```
### Predicted organ

The predicted organ comes from label transfer from the human fetal atlas. 
This is the reference developed by [Cao et al.](https://www.science.org/doi/10.1126/science.aba7721) provided by `Azimuth` for label transfer.
The reference contain cells from 15 organs including kidney from fetal samples.
The label transfer have been performed in the notebook `02a_fetal_full_reference_Cao_{sample_id}.Rmd`

@@

```{r fig.width=25, fig.height=6,  results='asis'}
tmp <- table(cell_type_df$sample_id, cell_type_df$organ)

knitr::kable(tmp)
```

```{r fig.width=25, fig.height=6}
do_Table_Heatmap(first_group = "sample_id", last_group = "organ")
```

### Predicted compartment

The predicted compartment is the result of the label transfer fron the human fetal kidney atlas: 
[Stewart et al.](https://www.science.org/doi/10.1126/science.aat5031) created a [human fetal kidney atlas](https://www.kidneycellatlas.org/). 
This reference contains only fetal kidney cells and has been precisely annotated by kidney experts. 
The label transfer have been performed in the notebook `02b_fetal_kidney_reference_Stewart_{sample_id}.Rmd`

@@

We think that the first level of annotation `fetal_kidney_predicted.compartment` can be useful in identifying non-cancer cells, at least the endothelial and immune cells.

```{r fig.width=25, fig.height=3, results='asis'}
tmp <- table(cell_type_df$sample_id, cell_type_df$compartment)

knitr::kable(tmp)
```

```{r fig.width=25, fig.height=3}

do_Table_Heatmap(first_group = "sample_id", last_group = "compartment")
```

### Evaluate annotations with marker genes

Here we evaluate with marker genes the identification of endothelial, immune cells sub-populations.

@@

```{r fig.height=3, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean(  feature = "ENSG00000110799",
                  group.by = "compartment")
```

@@

```{r fig.height=3, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean(  feature = "ENSG00000081237",
                  group.by = "compartment")
```
